---
title: "Final Presentation"
output: slidy_presentation
date: "2025-12-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggthemes)
library(scales)
library(ggrepel)
library(patchwork)
library(plotly)
StudentPerformanceFactors <- read.csv("StudentPerformanceFactors.csv")
student_exam_scores <- read.csv("student_exam_scores.csv")
AcademicPerformance <- read.csv("AcademicPerformance.csv")
## Cleaning the Exam_Score variable for outliers (scores less than 0 or greater than 100, because they are impossible in this case)
SPFcleaned <- StudentPerformanceFactors |> 
  mutate(Exam_Score = if_else(Exam_Score < 0 | Exam_Score > 100, NA, Exam_Score))
# Re-ordered motivation variable to go from low to medium to high
NEW_Motivation_Level <- c("Low", "Medium", "High")
SPFcleaned$Motivation_Level <- factor(SPFcleaned$Motivation_Level, levels = NEW_Motivation_Level)
print(levels(SPFcleaned$Motivation_Level))
## Cleaning the exam_score variable for outliers in the Student Exam Scores data set (scores less than 0 or greater than 55, because the grading scale is only that range)
SES2 <- student_exam_scores |> 
  mutate(exam_score = if_else(exam_score < 0 | exam_score > 55, NA, exam_score))
# Prepare data for plotting test scores by lunch status
plot_data <- AcademicPerformance |>
  select(lunch, math.score, reading.score, writing.score) |>
  pivot_longer(
    cols = c(math.score, reading.score, writing.score),
    names_to = "subject",
    values_to = "score"
  )
SPFcleaned <- StudentPerformanceFactors |> 
  mutate(Exam_Score = if_else(Exam_Score < 0 | Exam_Score > 100, NA, Exam_Score))
#Creating new variable that only includes students with learning disabilities                      
learning_disability = StudentPerformanceFactors |> filter(Learning_Disabilities == "Yes")

#Plot for Previous Score vs Exam Score for students with learning disability 
LD3 <- ggplot(learning_disability, aes(x = Previous_Scores, y = Exam_Score)) +
  geom_jitter() +
  geom_smooth() +
  labs(
    title = "Previous Score vs Exam Score",
    subtitle = "Student Performance Factors Dataset",
    x = "Previous Score",
    y = "Exam Score",
  )
ggplotly(LD3)
                    
                    
#Creating new variable that only includes students without learning disabilities            
no_disability = StudentPerformanceFactors |> filter(Learning_Disabilities == "No")

#Plot for Previous Score vs Exam Score for students without learning disability
NLD <- ggplot(no_disability, aes(x = Previous_Scores, y = Exam_Score)) +
  geom_jitter() +
  geom_smooth() +
  labs(
    title = "Previous Score vs Exam Score",
    subtitle = "Student Performance Factors Dataset",
    x = "Previous Score",
    y = "Exam Score",
  )
ggplotly(NLD)
```

## An Exploration of Lifestyle Attributes on Student Academic Performance

Names: Elsa Mathew-Lewis, Emily Sun, tAlec Anghel, Aoun Hakim

## The Number of Hours Spent Studying Has a Positive Linear Association with Exam Score - Those who study more appear to perform better
```{r}
SES2_HSbyES <- ggplot(SES2, aes(x = hours_studied, y = exam_score)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "The Impact of Hours Spent Studying on Grades",
    subtitle = "Exam Score Increases with Time Spent Studying",
    x = "Hours Studied per Week",
    y = "Exam Score (out of 55)",
    caption = "Source: Kaggle via Muhammad Shoaib"
  )

ggplotly(SES2_HSbyES)
```


## Student's Motivation Level Appears to Have Little to No Impact on Amount of Time Spent Studying and Exam Scores
```{r}
basicSPFggplot_facet <- ggplot(SPFcleaned, aes(x = Hours_Studied, y = Exam_Score)) +
  geom_point(aes(color = Motivation_Level, shape = Motivation_Level)) +
  geom_smooth() +
  facet_wrap(~ Motivation_Level) +  
  labs(
    title = "The Impact of Hours Spent Studying on Grades",
    subtitle = "Exam Score Increases with Time Spent Studying",
    x = "Hours Studied per Week",
    y = "Exam Score (out of 100)",
    color = "Motivation Level", shape = "Motivation Level",
    caption = "Source: Kaggle via Lai Ng"
  ) +
  scale_color_colorblind()

ggplotly(basicSPFggplot_facet)
```


## Students with standard lunch tend to have higher score on math, reading, and writing.
```{r}
ggplot(plot_data, aes(x = lunch, y = score, fill = lunch)) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(~ subject, nrow = 1,
             labeller = labeller(
               subject = c(
                 "math.score" = "Math",
                 "reading.score" = "Reading",
                 "writing.score" = "Writing"
               ))) +
  labs(
    title = "Test Scores by Lunch Status",
    subtitle = "students with standard lunch tend to have higher score on math, reading, and writing",
    x = "Lunch status",
    y = "Score",
    caption = "Source: Academic Performance dataset"
  ) +
  scale_x_discrete(labels = c("free/reduced" = "Free / Reduced",
                              "standard"    = "Standard")) +
  theme_minimal(base_size = 12)
```

## There is a Much Higher Number of Low to Medium Income Families with Children in Public School than Private School
```{r}
NEW_Family_Income <- c("Low", "Medium", "High")
SPFcleaned$Family_Income <- factor(SPFcleaned$Family_Income, levels = NEW_Family_Income)
p <- ggplot(SPFcleaned, aes(x = Family_Income, fill = School_Type)) +
  geom_bar(position = "dodge") +
  labs(title = "Family Income and School Type",
       x = "Family Income",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("lightblue", "pink"))

p_interactive_FamilyIncome_SchoolType <- ggplotly(p)
p_interactive_FamilyIncome_SchoolType
```

## Students without Learning Disabilities Have a Stronger Relationship between the Amount of Hours They Study and Their Exam Scores
```{r}
LD1NEW <- ggplot(SPFcleaned, aes(x = Hours_Studied, y = Exam_Score, color = Learning_Disabilities)) +
                      geom_jitter() +
                      geom_smooth() +
                      labs(
                        title = "Hours Studied vs Exam Score",
                        subtitle = "Student Performance Factors Dataset",
                        x = "Hours Studied",
                        y = "Exam Score",
                        color = "Learning Disability"
                      )
ggplotly(LD1NEW) 
```

## There Doesn't Appear to Be a Significant Difference In the Relationship Between Previous Exams Scores and Current Scores for Students with and without Learning Disabilities
```{r}
ldd = StudentPerformanceFactors |> mutate(ld = case_when(
                      Learning_Disabilities == "Yes" ~ "Has Learning Disability",
                      Learning_Disabilities == "No" ~ "No Learning Disability"
                    ))
LD_vs_NLD <- ggplot(ldd, aes(x = Previous_Scores, y = Exam_Score)) +
                      geom_jitter() +
                      facet_wrap(~ ld) +
                      geom_smooth() +
                      labs(
                        title = "Previous Score vs Exam Score",
                        subtitle = "Student Performance Factors Dataset",
                        x = "Previous Score",
                        y = "Exam Score"
                      )
ggplotly(LD_vs_NLD)
```

## Students Test Performance Indicates a Very Strong, Positive, and Linear Correlation between Writing Scores and Reading Scores 
```{r}
     ggplot(AcademicPerformance, aes(x = writing.score, y = reading.score)) +
                      geom_jitter() +
                      geom_smooth() +
                      labs(
                        title = "Writing vs Reading Scores",
                        subtitle = "Student Performance Dataset",
                        x = "Writing Score",
                        y = "Reading Score"
                      )
```

## Gender's Effect on Test Performance: Males Perform Better in Math on Average
```{r}
 genderandmath <- ggplot(AcademicPerformance, aes(x = gender, y = math.score)) +
                      geom_boxplot() +
                      labs(
                        title = "Gender effect on Math Scores",
                        subtitle = "Student Performance Dataset",
                        x = "Gender",
                        y = "Math Score"
                      )
                    ggplotly(genderandmath)
```

## Gender's Effect on Test Performance: Females Perform Better in Reading on Average
```{r}
genderandreading <- ggplot(AcademicPerformance, aes(x = gender, y = reading.score)) +
                      geom_boxplot() +
                      labs(
                        title = "Gender effect on Reading Scores",
                        subtitle = "Student Performance Dataset",
                        x = "Gender",
                        y = "Reading Score"
                      )
                    ggplotly(genderandreading)
```

## Lunch Status Effect on Reading Score: Reveals That Students with Lower Income Perform Worse
```{r}
 lunch_statusandreading <-ggplot(AcademicPerformance, aes(x = lunch, y = reading.score)) +
                      geom_boxplot() +
                      labs(
                        title = "Lunch status effect on Reading Scores",
                        subtitle = "Student Performance Dataset",
                        x = "Lunch",
                        y = "Reading Score"
                      )
                   ggplotly(lunch_statusandreading)
```

## Test Prep Effect on Math: Reveals that Students Who Do Test Prep Perform Better
```{r}
test_prepandmath <- ggplot(AcademicPerformance, aes(x = test.preparation.course, y = math.score)) +
                      geom_boxplot() +
                      labs(
                        title = "Test Preparation effect on Math Scores",
                        subtitle = "Student Performance Dataset",
                        x = "Test Preparation",
                        y = "Math Score"
                      )
                   ggplotly(test_prepandmath)
```

## Impact of Parental Involvement on Exam Scores: Reveals that Higher Parental Involvement Correlates with Higher Exam Scores
```{r}
parent = StudentPerformanceFactors |> mutate(Parental_Involvement = factor(Parental_Involvement, levels = c("Low", "Medium", "High")))
                   
      parent_involvementandscores <- ggplot(parent, aes(x = Parental_Involvement, y = Exam_Score)) +
                      geom_boxplot() +
                      labs(
                        title = r'(Parents' Involvement effect on Exam Scores)',
                        subtitle = "Student Performance Factors Dataset",
                        x = r"(Parents' Involvement)",
                        y = "Exam Score"
                      )
      ggplotly(parent_involvementandscores)
```